---
title: "Configuring EcoExtract for Your Data"
author: "EcoExtract Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Configuring EcoExtract for Your Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

EcoExtract is domain-agnostic and works with any JSON schema you define. This guide teaches you how to customize EcoExtract to extract exactly the data you need for your research domain.

## What You'll Learn

- How to create custom configuration files
- How to define your data schema (what fields to extract)
- How to write extraction prompts (how to extract those fields)
- How to use array fields for multi-valued data
- How to work with required vs optional fields
- How to test and iterate on your configuration

## Configuration Files

EcoExtract's extraction behavior is controlled by two key configuration files:

1. **`schema.json`** - Defines the structure of extracted data (what fields, what types)
2. **`extraction_prompt.md`** - Provides instructions to the LLM (how to extract, edge cases)

There's also an optional refinement prompt:

3. **`refinement_prompt.md`** - Instructions for the refinement step (validating and enhancing data)

And an optional metadata schema and prompt, for corpora that are not journal articles (see [Custom Document Metadata]):

4. **`metadata_schema.json`** - Document-level fields extracted once per document
5. **`metadata_prompt.md`** - Instructions for the metadata step

# Understanding the Schema

## Schema Basics

The schema is a JSON file that defines:

- What fields should be extracted
- What data types each field should have
- Which fields are required vs optional
- How to validate the extracted data
- Which fields uniquely identify records

## Required Structure

Your schema **must** follow this structure:

```json
{
  "$schema": "http://json-schema.org/draft-07/schema#",
  "title": "Your Domain Schema Title",
  "type": "object",
  "additionalProperties": false,
  "required": ["records"],
  "properties": {
    "records": {
      "type": "array",
      "description": "Array of extracted records",
      "items": {
        "type": "object",
        "additionalProperties": false,
        "required": ["field1", "field2", "optional_field"],
        "properties": {
          "field1": {
            "type": "string",
            "description": "A required string field"
          },
          "field2": {
            "type": "integer",
            "description": "A required integer field"
          },
          "optional_field": {
            "type": ["string", "null"],
            "description": "Optional - returns null when not available"
          }
        },
        "x-unique-fields": ["field1", "field2"]
      }
    }
  }
}
```

Key requirements:

- Top-level must have a `records` property (array of objects)
- Each field should have a `type` and `description`
- Every object must have `"additionalProperties": false`
- **All** properties must be listed in `required` -- use nullable types (`["string", "null"]`) for optional fields
- Use `x-unique-fields` to identify fields that define uniqueness (used for deduplication and accuracy)
- See [Cross-Provider Compatibility](#cross-provider-compatibility) below for full details

## Field Types

The schema supports standard JSON Schema types:

| JSON Type | SQLite Type | R Type | Example |
|-----------|-------------|--------|---------|
| `string` | TEXT | character | "Myotis lucifugus" |
| `integer` | INTEGER | integer | 2023 |
| `number` | REAL | numeric | 45.123 |
| `boolean` | BOOLEAN | logical | true |
| `array` | TEXT (JSON) | list | ["sentence 1", "sentence 2"] |
| `object` | TEXT (JSON) | list | {"lat": 45, "lon": -122} |

## Reserved System Field

**⚠️ IMPORTANT**: `record_id` is a reserved system field:

- **Format**: `{id fields}_{replicate}_r{N}` (e.g., `Smith_2020_1_r1`), built from the metadata fields named in the metadata schema's `x-record-id-fields`
- **Purpose**: Unique identifier for each record
- **Generated**: Automatically by the system
- **Do NOT include in your schema** - this is managed internally

# Creating Custom Configuration

## Step 1: Initialize Configuration

Create template configuration files in your project:

```{r init}
library(ecoextract)

# Creates ecoextract/ directory with template files
init_ecoextract()
```

This creates:

- `ecoextract/SCHEMA_GUIDE.md` - Read this first!
- `ecoextract/schema.json` - Template schema to customize
- `ecoextract/extraction_prompt.md` - Template prompt to customize

## Step 2: Edit the Schema

Open `ecoextract/schema.json` and customize for your domain.

### Example 1: Simple Schema (Disease Outbreaks)

```json
{
  "$schema": "http://json-schema.org/draft-07/schema#",
  "title": "Disease Outbreak Schema",
  "description": "Schema for extracting disease outbreak data from epidemiological literature",
  "type": "object",
  "additionalProperties": false,
  "required": ["records"],
  "properties": {
    "records": {
      "type": "array",
      "description": "Array of disease outbreak records",
      "items": {
        "type": "object",
        "additionalProperties": false,
        "required": [
          "disease_name",
          "location",
          "year",
          "cases",
          "deaths",
          "all_supporting_source_sentences"
        ],
        "properties": {
          "disease_name": {
            "type": "string",
            "description": "Name of the disease (e.g., 'COVID-19', 'Influenza A(H1N1)')"
          },
          "location": {
            "type": "string",
            "description": "Geographic location of outbreak (city, region, or country)"
          },
          "year": {
            "type": "integer",
            "description": "Year the outbreak occurred"
          },
          "cases": {
            "type": ["integer", "null"],
            "description": "Total number of cases reported (null if not stated)"
          },
          "deaths": {
            "type": ["integer", "null"],
            "description": "Total number of deaths reported (null if not stated)"
          },
          "all_supporting_source_sentences": {
            "type": "array",
            "description": "All sentences from the document that support this record",
            "items": {
              "type": "string",
              "description": "Individual supporting sentence"
            }
          }
        },
        "x-unique-fields": ["disease_name", "location", "year"]
      }
    }
  }
}
```

### Example 2: Complex Schema with Arrays (Host-Pathogen Interactions)

```json
{
  "$schema": "http://json-schema.org/draft-07/schema#",
  "title": "Host-Pathogen Interaction Schema",
  "type": "object",
  "additionalProperties": false,
  "required": ["records"],
  "properties": {
    "records": {
      "type": "array",
      "items": {
        "type": "object",
        "additionalProperties": false,
        "required": [
          "Pathogen_Name",
          "Host_Name",
          "Detection_Method",
          "Sample_type",
          "Location",
          "all_supporting_source_sentences",
          "Confidence_Score"
        ],
        "properties": {
          "Pathogen_Name": {
            "type": "string",
            "description": "Scientific name of the pathogen"
          },
          "Host_Name": {
            "type": "string",
            "description": "Scientific name of the host organism"
          },
          "Detection_Method": {
            "type": "array",
            "description": "Methods used to detect the pathogen (can be multiple)",
            "items": {
              "type": "string",
              "description": "Individual detection method (e.g., 'PCR', 'culture', 'serology')"
            }
          },
          "Sample_type": {
            "type": "array",
            "description": "Types of samples collected (empty array if not mentioned)",
            "items": {
              "type": "string",
              "description": "Individual sample type (e.g., 'blood', 'tissue', 'fecal swab')"
            }
          },
          "Location": {
            "type": ["string", "null"],
            "description": "Geographic location where interaction was documented"
          },
          "all_supporting_source_sentences": {
            "type": "array",
            "description": "All sentences from document supporting this record",
            "items": {
              "type": "string"
            }
          },
          "Confidence_Score": {
            "type": ["integer", "null"],
            "description": "Confidence in extraction accuracy (1-5 scale)"
          }
        },
        "x-unique-fields": ["Pathogen_Name", "Host_Name"]
      }
    }
  }
}
```

## Step 3: Edit the Extraction Prompt

Open `ecoextract/extraction_prompt.md` and customize the instructions for your domain.

### Prompt Structure

A good extraction prompt should include:

1. **Task description** - What you're extracting and why
2. **Field definitions** - Clear explanation of each field
3. **Extraction rules** - When to create separate records
4. **Edge case handling** - What to do with ambiguous cases
5. **Quality guidelines** - Standards for completeness and accuracy

### Example Prompt (Disease Outbreaks)

```markdown
# Task

Extract all disease outbreak events mentioned in this epidemiological document.
Each outbreak should be a separate record.

# Field Definitions

- **disease_name**: The specific disease or pathogen name. Use the most specific
  name mentioned (e.g., "Influenza A(H1N1)" not just "Influenza").

- **location**: Geographic location where outbreak occurred. Use the most specific
  location mentioned (city > region > country).

- **year**: Year the outbreak occurred (4-digit year).

- **cases**: Total number of confirmed or suspected cases. Extract only if explicitly
  stated.

- **deaths**: Total number of deaths attributed to the outbreak. Extract only if
  explicitly stated.

- **all_supporting_source_sentences**: Include every sentence that provides
  information about this outbreak. Include the full sentence verbatim.

# Extraction Rules

## One Record Per Outbreak

Create separate records for:
- Different diseases in the same location
- Same disease in different locations
- Same disease in same location but different time periods

## Handling Uncertainty

- If year is not explicitly stated, try to infer from context
- If location is ambiguous, use the most likely interpretation
- If case/death counts are ranges, use the midpoint
- Mark low-confidence extractions with Confidence_Score = 2 or 3

## Multiple Mentions

If an outbreak is mentioned multiple times, create ONE record and include
all supporting sentences.

# Quality Standards

- Prioritize completeness: extract all outbreaks mentioned
- Include all supporting sentences for each record
- Be precise with location and disease names
- Don't hallucinate: only extract what's explicitly stated or clearly implied
```

## Step 4: Test Your Configuration

```{r test-config}
# Process a test document with your custom configuration
results <- process_documents(
  pdf_path = "test_paper.pdf",
  db_conn = "test.db"
)

# Retrieve and inspect results
records <- get_records(db_conn = "test.db")
View(records)

# Check what was extracted
names(records)  # Should match your schema fields
str(records)    # Check data types
```

## Step 5: Iterate and Refine

```{r iterate}
# Common iterations:
# 1. Adjust field descriptions in schema
# 2. Add edge case handling in prompt
# 3. Reprocess with force flag
# 4. Review results in ecoreview app

# Force reprocess after changing schema/prompt
results <- process_documents(
  pdf_path = "test_paper.pdf",
  db_conn = "test.db",
  force_reprocess_extraction = TRUE
)
```

# Custom Document Metadata

Before extraction, the metadata step pulls document-level fields into the `documents` table. The package default extracts bibliographic metadata for journal articles (title, authors, publication year, DOI, journal, ...). For other corpora, such as shipment declarations or permits, supply your own metadata schema and prompt:

- **`ecoextract/metadata_schema.json`** replaces the default metadata schema
- **`ecoextract/metadata_prompt.md`** replaces the default metadata prompt

To use files elsewhere, pass them explicitly; explicit files take priority over `ecoextract/`:

```r
process_documents("pdfs/",
                  metadata_schema_file = "config/metadata_schema.json",
                  metadata_prompt_file = "config/metadata_prompt.md")
```

If your documents have no useful document-level metadata, skip the step with `run_metadata = FALSE`. This saves one LLM call per document. Pair it with a metadata schema whose `x-record-id-fields` is `["document_id"]` or `["file_name"]`, so record IDs don't depend on metadata values.

The metadata schema wraps its fields in a single object under `properties` and names the fields that form record IDs in `x-record-id-fields`. The object can have any name; the package default calls it `publication_metadata`, and the example below uses `shipment_metadata`. The model sees this name in the structured output schema, so pick one that describes your documents:

```json
{
  "$schema": "http://json-schema.org/draft-07/schema#",
  "type": "object",
  "additionalProperties": false,
  "properties": {
    "shipment_metadata": {
      "type": "object",
      "description": "Shipment declaration metadata",
      "additionalProperties": false,
      "x-record-id-fields": ["control_number"],
      "required": ["control_number", "port_of_entry", "filing_date"],
      "properties": {
        "control_number": {
          "type": ["string", "null"],
          "description": "Declaration control number printed on Form 3-177"
        },
        "port_of_entry": {
          "type": ["string", "null"],
          "description": "Port where the shipment entered"
        },
        "filing_date": {
          "type": ["string", "null"],
          "description": "Date the declaration was filed (YYYY-MM-DD)"
        }
      }
    }
  },
  "required": ["shipment_metadata"]
}
```

The `documents` table gets one column per metadata field, created automatically. Array and object fields are stored as JSON text. Only the fields in the schema in use become columns: a database created with a custom metadata schema has no bibliographic columns (title, authors, journal, and so on), and `export_db()` exports the schema's fields. `export_bibtex()` needs the bibliographic fields, so it only works with the default metadata schema.

`x-record-id-fields` can name metadata fields, `document_id`, or `file_name` (used without its extension). With the schema above, records get IDs like `2024123456_1_r1`; with `["document_id"]` they get `42_1_r1`. Choose fields with short values: each value is stripped to letters and digits and used in full.

**Keep metadata schemas within model limits.** Older models enforce tight structured-output limits and report them only as `Schema is too complex`. Claude Sonnet 4.6, for example, accepts at most 12 optional fields and 16 nullable (union-typed) fields. List every field in `required` and use nullable types (`["string", "null"]`) for values that may be absent, as the example does, and keep nullable fields to 16 or fewer.

# Working with Array Fields

Array fields let you capture multi-valued data like:
- Multiple detection methods per observation
- Multiple sample types
- Multiple supporting sentences
- Multiple geographic locations

## Defining Array Fields

```json
{
  "detection_methods": {
    "type": "array",
    "description": "All methods used to detect the organism",
    "items": {
      "type": "string",
      "description": "Individual detection method (e.g., 'PCR', 'culture', 'ELISA')"
    }
  }
}
```

## Instructing the LLM on Arrays

In your extraction prompt, clearly explain when to use arrays:

```markdown
## detection_methods (array)

List ALL methods mentioned for detecting this organism. Each method should be a
separate array element.

Examples:
- If text says "detected by PCR and culture" → ["PCR", "culture"]
- If text says "PCR-positive samples" → ["PCR"]
- If text says "antibodies detected via ELISA" → ["ELISA"]

Common methods: PCR, RT-PCR, qPCR, culture, serology, ELISA, Western blot,
microscopy, sequencing, antigen detection
```

## Retrieving Array Data in R

```{r arrays}
library(dplyr)
library(tidyr)

# Get records with array fields
records <- get_records(db_conn = "records.db")

# Array fields are stored as lists
class(records$detection_methods)  # "list"

# Unnest to see individual values
records |>
  unnest(detection_methods) |>
  select(record_id, detection_methods)

# Count frequency of each method
records |>
  unnest(detection_methods) |>
  count(detection_methods, sort = TRUE)
```

## Best Practices for Arrays

1. **Be explicit in prompts** - Show examples of single vs multiple values
2. **Use consistent terminology** - Define a controlled vocabulary when possible
3. **Always include descriptions** - Help the LLM understand what belongs in the array
4. **Test with edge cases** - Papers with 1 value, 5 values, 0 values

# Structural vs Data Requirements

The `required` array in JSON Schema serves two purposes that are important to distinguish:

## Structural Requirement (for OpenAI)

OpenAI's structured outputs API requires **every** property to be listed in `required`. This is a structural constraint -- it ensures the model always returns all fields in its response. It does NOT mean every field must contain meaningful data.

## Data Requirements (for your domain)

The **type** of each field controls whether it must contain data:

| Pattern | Meaning | LLM returns when empty | Database |
|---------|---------|------------------------|----------|
| `"type": "string"` | Must have data | Always a string value | NOT NULL |
| `"type": ["string", "null"]` | Data optional | `null` | NULL allowed |
| `"type": "array"` | May be empty | `[]` (empty array) | `"[]"` |

## Example

```json
{
  "items": {
    "type": "object",
    "additionalProperties": false,
    "required": [
      "pathogen_name",
      "host_name",
      "location",
      "detection_methods",
      "all_supporting_source_sentences"
    ],
    "properties": {
      "pathogen_name": { "type": "string", "description": "Always present" },
      "host_name": { "type": "string", "description": "Always present" },
      "location": { "type": ["string", "null"], "description": "Null if not mentioned" },
      "detection_methods": { "type": "array", "items": { "type": "string" }, "description": "Empty array if not mentioned" },
      "all_supporting_source_sentences": { "type": "array", "items": { "type": "string" }, "description": "Supporting evidence" }
    }
  }
}
```

All five fields are structurally required (in `required`), but only `pathogen_name` and `host_name` must have data. `location` can be `null`, and `detection_methods` can be `[]`.

**Important**: Do NOT use `["array", "null"]` for optional arrays. Use plain `"array"` and instruct the model to return `[]` when no data is available.

## Impact on Downstream Processing

- **Accuracy calculation** - Edits to non-nullable fields are "major edits"
- **Deduplication** - Fields in `x-unique-fields` should generally be non-nullable
- **Database constraints** - Non-nullable fields get `NOT NULL` constraints; nullable fields allow NULL

## Guidelines

**Non-nullable (truly required):**
- Core identifying information (organism names)
- Evidence fields (`all_supporting_source_sentences`)
- Fields needed for deduplication

**Nullable (optional):**
- Descriptive details that may not always be present
- Quantitative measurements (counts, concentrations)
- Contextual information (dates, locations, page numbers)

**Array fields (may be empty):**
- Multi-valued fields like detection methods, sample types, vector names
- Use the field description to indicate when empty arrays are acceptable

# Unique Fields and Deduplication

## Defining Unique Fields

Use `x-unique-fields` to specify which fields define record uniqueness:

```json
{
  "items": {
    "type": "object",
    "additionalProperties": false,
    "required": ["pathogen_name", "host_name", "location", "sample_date"],
    "properties": {
      "pathogen_name": { "type": "string", "description": "Pathogen name" },
      "host_name": { "type": "string", "description": "Host name" },
      "location": { "type": ["string", "null"], "description": "Location" },
      "sample_date": { "type": ["string", "null"], "description": "Date" }
    },
    "x-unique-fields": ["pathogen_name", "host_name", "location"]
  }
}
```

## How Unique Fields Are Used

1. **Deduplication** - Records with same unique field values are considered duplicates
2. **Accuracy metrics** - Edits to unique fields are classified as "major edits"
3. **Review workflow** - Helps identify truly distinct records

## Choosing Unique Fields

Good unique field combinations:
- **Species interactions**: species1 + species2 + interaction_type
- **Disease outbreaks**: disease + location + year
- **Chemical measurements**: chemical + sample_location + date
- **Observations**: organism + location + date + observer

# Cross-Provider Compatibility

ecoextract supports multiple LLM providers (Anthropic Claude, OpenAI GPT, Google Gemini, Mistral). Schemas must follow [OpenAI's structured output requirements](https://platform.openai.com/docs/guides/structured-outputs/supported-schemas) to work across all providers:

1. **`additionalProperties: false`** on every object definition
2. **All properties listed in `required`** -- every field in `properties` must also appear in `required`
3. **Nullable types for optional fields** -- use `"type": ["string", "null"]` instead of omitting from `required`. The model returns `null` when no data is available
4. **No `minimum`/`maximum` constraints** -- put range info in the `description` instead (e.g., `"description": "Confidence score (1-5 scale)"`)
5. **No `$ref`** -- inline all definitions
6. **Max 100 properties, 5 nesting levels**

These rules are safe for all providers -- Claude, Gemini, and Mistral handle them without issues. Note: `additionalProperties` is automatically stripped for Gemini, which does not support it.

For full details, see [OpenAI's Structured Outputs documentation](https://platform.openai.com/docs/guides/structured-outputs/supported-schemas).

# Advanced Configuration

## Configuration Priority

EcoExtract looks for configuration files in this order:

1. **Explicit parameter** - Files you specify directly in function calls
2. **Project ecoextract/ directory** - `ecoextract/schema.json`, etc.
3. **Working directory with prefix** - `ecoextract_schema.json`, etc.
4. **Package defaults** - Built-in example schema

```{r config-priority}
# Uses ecoextract/schema.json automatically
process_documents("pdfs/", "records.db")

# Specify custom path explicitly
process_documents(
  pdf_path = "pdfs/",
  db_conn = "records.db",
  schema_file = "custom/my_schema.json",
  extraction_prompt_file = "custom/my_prompt.md"
)
```

## Refinement Prompts

The refinement step validates and enhances extracted data. Customize the refinement prompt for domain-specific validation:

```{r refinement}
# Run with custom refinement prompt
process_documents(
  pdf_path = "pdfs/",
  db_conn = "records.db",
  run_refinement = TRUE,
  refinement_prompt_file = "ecoextract/refinement_prompt.md"
)
```

A good refinement prompt should:
- Validate field values against known standards
- Check for logical consistency
- Enhance incomplete records using context
- Resolve ambiguities from extraction step

## Multiple Schemas for Different Projects

```{r multiple-schemas}
# Project 1: Disease outbreaks
process_documents(
  "disease_pdfs/",
  "disease.db",
  schema_file = "schemas/disease_schema.json",
  extraction_prompt_file = "prompts/disease_prompt.md"
)

# Project 2: Species interactions
process_documents(
  "species_pdfs/",
  "species.db",
  schema_file = "schemas/species_schema.json",
  extraction_prompt_file = "prompts/species_prompt.md"
)
```

# Complete Example: Building a Custom Schema

Let's walk through creating a schema for extracting plant-pollinator interactions:

## 1. Define Your Domain

What information do we need?
- Plant species (scientific name)
- Pollinator species (scientific name)
- Interaction type (pollination, nectar robbing, etc.)
- Location (where observed)
- Date/season (when observed)
- Supporting evidence

## 2. Create the Schema

```{r example-schema, eval=FALSE}
# Save as ecoextract/schema.json
{
  "$schema": "http://json-schema.org/draft-07/schema#",
  "title": "Plant-Pollinator Interaction Schema",
  "type": "object",
  "additionalProperties": false,
  "required": ["records"],
  "properties": {
    "records": {
      "type": "array",
      "items": {
        "type": "object",
        "additionalProperties": false,
        "required": [
          "plant_species",
          "pollinator_species",
          "interaction_type",
          "location",
          "season",
          "all_supporting_source_sentences"
        ],
        "properties": {
          "plant_species": {
            "type": "string",
            "description": "Scientific name of plant species"
          },
          "pollinator_species": {
            "type": "string",
            "description": "Scientific name of pollinator species"
          },
          "interaction_type": {
            "type": "array",
            "items": {
              "type": "string",
              "description": "Type of interaction: pollination, nectar_robbing, etc."
            }
          },
          "location": {
            "type": ["string", "null"],
            "description": "Geographic location of observation"
          },
          "season": {
            "type": ["string", "null"],
            "description": "Season or months when interaction observed"
          },
          "all_supporting_source_sentences": {
            "type": "array",
            "items": {
              "type": "string"
            }
          }
        },
        "x-unique-fields": ["plant_species", "pollinator_species", "location"]
      }
    }
  }
}
```

## 3. Write the Prompt

```{r example-prompt, eval=FALSE}
# Save as ecoextract/extraction_prompt.md

# Task
Extract all plant-pollinator interactions reported in this document.

# Field Definitions

**plant_species**: Scientific name of the plant (genus + species).

**pollinator_species**: Scientific name of the pollinator. Use the most specific
taxonomic level available.

**interaction_type**: Type(s) of interaction. Common types:
- pollination (legitimate pollination)
- nectar_robbing (accessing nectar without pollinating)
- pollen_collecting (collecting pollen but not pollinating)

**location**: Geographic location. Use most specific available (site > region > country).

**season**: When the interaction was observed (season, months, or dates).

**all_supporting_source_sentences**: All sentences providing information about
this interaction.

# Extraction Rules

Create one record per unique plant-pollinator-location combination.

If the same pair is observed in multiple seasons at the same location, create
separate records.

# Examples

"Bombus terrestris was observed visiting Trifolium repens flowers in meadows
near Oxford during June and July."

→ One record:
- plant_species: "Trifolium repens"
- pollinator_species: "Bombus terrestris"
- interaction_type: ["pollination"]
- location: "meadows near Oxford"
- season: "June and July"
```

## 4. Process and Test

```{r example-process}
# Process test documents
process_documents(
  pdf_path = "pollinator_papers/",
  db_conn = "pollinators.db"
)

# Review results
records <- get_records(db_conn = "pollinators.db")

# Check extraction quality
library(ecoreview)
run_review_app("pollinators.db")

# Calculate accuracy after review
accuracy <- calculate_accuracy("pollinators.db")
```

# Troubleshooting

## Schema Validation Errors

```{r schema-errors}
# Validate JSON syntax
jsonlite::validate("ecoextract/schema.json")

# Read and inspect schema
schema <- jsonlite::read_json("ecoextract/schema.json")
names(schema$properties)  # Should include "records"

# Test with default schema first
process_documents("test.pdf", "test.db", schema_file = NULL)
```

## Unexplained Schema Errors

If the API rejects a schema with `Schema is too complex` or another HTTP 400 that points at the schema, **try a newer model first**. These errors are usually about the model, not the schema: older models enforce much tighter structured-output limits than current ones. The package's own default metadata schema, before it was restructured, failed on `anthropic/claude-sonnet-4-6` and `anthropic/claude-haiku-4-5` and passed on `anthropic/claude-sonnet-5` and `anthropic/claude-opus-5`, with nothing about the schema needing to change.

These errors are easy to misread:

- The message names the schema and gives no field count, limit, or offending field.
- Anthropic does not publish these limits, and they have tightened over time without a changelog entry, so a schema that used to work can start failing.
- `process_documents()` tries every model in `model` in turn. If all of them are older models, every attempt fails with the same error and the failure looks like it comes from the schema.

If you need to stay on an older model, these are the limits measured on `anthropic/claude-sonnet-4-6` for a single flat object (September 2026):

| Field style | Limit |
|---|---|
| Optional (omitted from `required`) | 12 |
| Required and nullable (`["string", "null"]`) | 16 |
| Required, single type | no limit reached at 40 |

Optional and nullable fields are the expensive constructs. The limits also share a combined budget, with an optional field costing roughly three to four times a required one (12 optional plus 28 required fails, though each passes alone). The union limit is the only one the API states outright:

```
Schemas contains too many parameters with union types (18 parameters with
type arrays or anyOf). ... (limit: 16 parameters with unions).
```

To fit a schema within these limits, list every field in `required`, use nullable types only for values that may be absent, and keep nullable fields to 16 or fewer.

## LLM Not Following Instructions

If the LLM isn't extracting correctly:

1. **Make prompts more explicit** - Add examples and edge cases
2. **Simplify field descriptions** - Use clear, unambiguous language
3. **Check required fields** - Too many required fields can cause failures
4. **Test incrementally** - Start with a simple schema, add complexity gradually

## Array Fields Not Working

```{r array-issues}
# Check if arrays are being recognized
records <- get_records()
class(records$your_array_field)  # Should be "list"

# If it's character, arrays aren't being parsed correctly
# Check schema has proper array definition:
# "type": "array", "items": {"type": "string"}
```

# Next Steps

- **Process your documents** with your custom schema
- **Launch the review app** to validate extraction quality
- **Iterate on configuration** based on results
- **Calculate accuracy** to measure performance

## Additional Resources

- [Complete Workflow Guide](ecoextract-workflow.html) - Full workflow reference
- [ACCURACY.md](../ACCURACY.md) - Understanding accuracy metrics
- [SCHEMA_GUIDE.md](../inst/extdata/SCHEMA_GUIDE.md) - Detailed schema specification
- [OpenAI Structured Outputs](https://platform.openai.com/docs/guides/structured-outputs/supported-schemas) - Schema requirements for cross-provider compatibility
