---
title: "Complete Workflow: From PDF to Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow: From PDF to Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This vignette demonstrates the complete workflow for extracting ecological interaction data from scientific literature PDFs using `ecoextract`.

The pipeline consists of four main steps:

1. **OCR Processing** - Convert PDF to markdown text
2. **Document Audit** - Quality check the OCR output
3. **Extraction** - Extract structured interaction data using LLM
4. **Refinement** - Enhance and verify extracted data

## Installation

```{r setup}
library(ecoextract)

# Set your Anthropic API key
Sys.setenv(ANTHROPIC_API_KEY = "your-api-key-here")
```

## Quick Start: Process a Single File or Folder

The simplest way to process documents is using `process_documents()`:

```{r quickstart}
# Process a single PDF
process_documents(
  pdf_path = "my_paper.pdf",
  db_path = "interactions.db"
)

# Process all PDFs in a folder
process_documents(
  pdf_path = "pdfs/",
  db_path = "interactions.db"
)
```

This automatically handles:
- OCR processing
- OCR quality audit
- Data extraction with Claude
- Refinement pass
- Saving to SQLite database

## Customizing the Workflow

### Step 1: Initialize Custom Configuration

Create customizable prompt and schema files in your project:

```{r init-config}
# Creates ecoextract/ directory with template files
init_ecoextract()

# This creates:
# - ecoextract/schema.json
# - ecoextract/extraction_prompt.md
# - ecoextract/refinement_prompt.md
```

### Step 2: Customize for Your Domain

Edit the files in `ecoextract/` to customize for your research domain:

**ecoextract/schema.json** - Define what data to extract:
```json
{
  "type": "object",
  "properties": {
    "interactions": {
      "type": "array",
      "items": {
        "type": "object",
        "properties": {
          "species_a": {"type": "string"},
          "species_b": {"type": "string"},
          "interaction_type": {"type": "string"},
          "location": {"type": "string"}
        }
      }
    }
  }
}
```

**ecoextract/extraction_prompt.md** - Customize extraction instructions

**ecoextract/refinement_prompt.md** - Customize refinement instructions

### Step 3: Process with Custom Configuration

The package automatically detects and uses files in `ecoextract/`:

```{r custom-process}
# Automatically uses ecoextract/schema.json, ecoextract/extraction_prompt.md, etc.
process_documents("pdfs/", "interactions.db")

# Or specify custom files explicitly
process_documents(
  pdf_path = "pdfs/",
  db_path = "interactions.db",
  schema_file = "my_custom/schema.json",
  extraction_prompt_file = "my_custom/extraction.md"
)
```

## Understanding the Workflow Steps

The `process_documents()` function handles all steps automatically, but you can understand what happens at each stage:

### Step 1: OCR Processing

Converts PDF to markdown text using OCR. The package uses external OCR libraries and stores the result in the database.

### Step 2: Document Audit

Quality check of the OCR output to identify potential issues (missing tables, unclear text, etc.).

### Step 3: Extraction

Uses LLM (Claude) to extract structured interaction data from the markdown text according to your schema.

### Step 4: Refinement

Enhances extracted data by filling missing information, improving evidence quality, and cross-referencing with document tables.

All results are automatically saved to the SQLite database with proper record IDs.

## Working with the Database

### Query Records

```{r query}
library(DBI)
library(dplyr)

# Connect to database
db_conn <- dbConnect(RSQLite::SQLite(), "interactions.db")

# Get all records
records <- tbl(db_conn, "records") %>%
  collect()

# Filter by species
bat_records <- tbl(db_conn, "records") %>%
  filter(grepl("Myotis", bat_species_scientific_name)) %>%
  collect()

# Join with document info
full_data <- tbl(db_conn, "records") %>%
  left_join(tbl(db_conn, "documents"), by = "document_id") %>%
  collect()

dbDisconnect(db_conn)
```

### Export Data

```{r export}
# Export to CSV
library(readr)
write_csv(records, "extracted_records.csv")
```

## Configuration Priority

The package looks for configuration files in this order:

1. **Explicit parameter** - Files you specify directly
   ```r
   extract_records(schema_file = "path/to/schema.json")
   ```

2. **Project ecoextract/ directory** - Project-specific configs
   ```
   your_project/
   ├── ecoextract/
   │   ├── schema.json
   │   ├── extraction_prompt.md
   │   └── refinement_prompt.md
   └── analysis.R
   ```

3. **Working directory with prefix** - Alternative project location
   ```
   your_project/
   ├── ecoextract_schema.json
   ├── ecoextract_extraction_prompt.md
   └── analysis.R
   ```

4. **Package defaults** - Built-in defaults for bat-plant interactions

## Best Practices

### 1. Start with Defaults

```{r defaults}
# Use package defaults first to understand the system
process_documents("test.pdf", "test.db")
```

### 2. Customize Incrementally

```{r incremental}
# Initialize config
init_ecoextract()

# Edit one file at a time
# - Start with schema.json
# - Then customize extraction_prompt.md
# - Finally tune refinement_prompt.md
```

### 3. Version Control Your Configs

```{r version-control, eval=FALSE}
# Add to git
git add ecoextract/
git commit -m "Add custom extraction schema for bird interactions"
```

### 4. Test on Small Batches

```{r test-batch}
# Test on 2-3 papers first
test_files <- list.files("pdfs", pattern = "\\.pdf$", full.names = TRUE)[1:3]

for (file in test_files) {
  process_documents(file, "test.db")
}

# Review results before processing entire corpus
```

### 5. Monitor Token Usage

```{r monitor}
# Extraction and refinement use Claude API
# Monitor your usage at: https://console.anthropic.com/

# Typical usage per paper:
# - Extraction: ~5,000-10,000 tokens
# - Refinement: ~3,000-5,000 tokens
```

## Troubleshooting

### OCR Not Available

```{r ocr-error, eval=FALSE}
# Error: OCR functionality requires the 'ohseer' package

# Solution: Install ohseer or provide markdown directly
extract_records(document_content = "your markdown text")
```

### API Key Not Found

```{r api-key-error, eval=FALSE}
# Error: API key not found

# Solution: Set environment variable
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
# The API key will be automatically detected by ellmer

# Or load from .env file
library(ecoextract)
load_env()
```

### Schema Validation Errors

```{r schema-error, eval=FALSE}
# If extraction fails, check your schema

# 1. Validate JSON syntax
jsonlite::validate("ecoextract/schema.json")

# 2. Test with default schema first
extract_records(document_content = text, schema_file = NULL)
```

## Next Steps

- Read the schema documentation: `?load_schema`
- Explore prompt customization: `view_prompt("extraction_prompt")`
- Learn about database structure: `?init_ecoextract_database`
